---
title: "Untitled"
author: "Maddee Rubenson (FlowWest)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(sf)
library(tidyverse)
library(leaflet)
library(terra)

```

## Objective: 

Evaluate how the removal of levees and dikes along the Sprague and Williamson Rivers may influence hydrologic connectivity, floodplain inundation patterns, and potential implications for water availability and agricultural lands.


```{r include=FALSE}
klamath_basin <- st_read('klamath_basin_outline/R8_FAC_Klamath_Basin_WFL1.shp')
```
### Read in Consumptive Use/Field data 

This data comes from Desert Research Institute: https://www.dri.edu/project/owrd-et/. The study provides a long-term record of monthly and annual satellite-based actual crop ET and consumptive use of irrigation water, potential crop ET and irrigation water requirements, applied irrigation water, and open water evaporation estimates for all major water bodies within the Oregon hydrographic area. 


```{r eval=FALSE, include=FALSE}
or_field <- st_layers(here::here("data-raw", "cropET", "Appendix_6a_Field_and_HUC_Consumptive_Use_Geodatabase", "or_field_geopackage.gpkg"))
layers <- unique(or_field$name)

# Mode of CDL classification
crop <- st_read(
  here::here("data-raw", "cropET", "Appendix_6a_Field_and_HUC_Consumptive_Use_Geodatabase", "or_field_geopackage.gpkg"),
  layer = "CROP"
) |> 
  st_transform(st_crs(klamath_basin))

# Monthly ensemble actual ET volume in acre-feet
et_volume <- st_read(
  here::here("data-raw", "cropET", "Appendix_6a_Field_and_HUC_Consumptive_Use_Geodatabase", "or_field_geopackage.gpkg"),
  layer = "ET_VOLUME"
) |> 
  st_transform(st_crs(klamath_basin))

# Monthly irrigation consumptive use volume in acre-feet (OpenET ensemble actual ET volume minus ET Demands Prz volume)
irr_cu_volume <- st_read(
  here::here("data-raw", "cropET", "Appendix_6a_Field_and_HUC_Consumptive_Use_Geodatabase", "or_field_geopackage.gpkg"),
  layer = "IRR_CU_VOLUME"
) |> 
  st_transform(st_crs(klamath_basin))

crop_crosswalk <- readxl::read_excel(here::here("data-raw", "cropET", "Appendix_4_CDL_ETDemands_crosswalk.xlsx"), skip = 1) |> janitor::clean_names()

klamath_crops <- sf::st_intersection(crop, klamath_basin)
klamath_et <- sf::st_intersection(et_volume, klamath_basin)
klamath_irr_vol <- sf::st_intersection(irr_cu_volume, klamath_basin)

saveRDS(klamath_crops, here::here("data-raw", "klamath_crops.rds"))
saveRDS(klamath_et, here::here("data-raw", "klamath_et.rds"))
saveRDS(klamath_irr_vol, here::here("data-raw", "klamath_irr_vol.rds"))
```

```{r}
klamath_crops <- read_rds(here::here("data-raw", "klamath_crops.rds"))
klamath_et <- read_rds(here::here("data-raw", "klamath_et.rds"))
klamath_irr_vol <- read_rds(here::here("data-raw", "klamath_irr_vol.rds"))
```

```{r include=FALSE}
unique(klamath_crops$ITYPE)

unique(klamath_crops$CROP_2022)

```

### Levees

**Data source: ** https://resilience.climate.gov/maps/87acff1ba86c40098b59472292de3d11/about

This data source was not included due to its limited spatial granularity within the Klamath Basin and the absence of coverage in the Upper Basin.


```{r echo=TRUE}

url <- "https://services2.arcgis.com/FiaPA4ga0iQKduv3/ArcGIS/rest/services/NLD2_PUBLIC_v1/FeatureServer/0/query?where=1%3D1&outFields=*&f=geojson"

nld <- st_read(url)

nld_klamath <- nld |> 
  st_transform(st_crs(klamath_basin)) |> 
  sf::st_intersection(klamath_basin) |> 
  janitor::clean_names()

ggplot() + 
  geom_sf(data = klamath_basin) + 
  geom_sf(data = nld_klamath) +
  theme_minimal()
```
```{r eval=FALSE, include=FALSE}
leaflet() |>
  addProviderTiles(providers$CartoDB.Positron, group = "Light") |>
  addProviderTiles(providers$Esri.WorldImagery, group = "Imagery") |>
  addPolygons(
    data = klamath_crops |> st_transform(4326),
    fillColor = "green",
    fillOpacity = 0.4,
    color = "darkgreen",
    weight = 1,
    group = "Crops"
  ) |>
  addPolylines(
    data = rivermile::all_klamath_rivers_line,
    color = "blue",
    weight = 1.5,
    opacity = 0.5,  
    label = ~river, 
    labelOptions = labelOptions(
      noHide = TRUE,         
      direction = "auto",
      textsize = "11px",
      style = list(
        "color" = "navy",
        "font-weight" = "bold"
      )
    ),
    group = "Rivers"
  ) |>
  addLayersControl(
    baseGroups = c("Light", "Imagery"),
    overlayGroups = c("Crops", "Rivers"),
    options = layersControlOptions(collapsed = FALSE)
  )
  

```

### DEM

The DEM was downloaded using the `elevatr` package. Elevation values are in meters. We also have access to higher-resolution DEM and bathymetric datasets should further analysis be warranted. The results presented here are exploratory and currently just for the Sprague River. 

```{r eval=FALSE}
library(elevatr)
library(terra)

aoi <- klamath_basin

dem <- get_elev_raster(
  locations = aoi,
  z = 12,          # higher = finer (~10–30 m typical)
  clip = "locations"
)

dem_rast <- rast(dem)
plot(dem_rast)

dem_wrap <- terra::wrap(dem_rast)
saveRDS(dem_wrap, here::here("data-raw", "DEM", "dem_wrap.rds"))


saveRDS(dem, here::here("data-raw", "DEM", "dem.rds"))
saveRDS(dem_rast, here::here("data-raw", "DEM", "dem_rast.rds"))

```
## Map Floodplain Inundation 

The bathtub approach is a simple DEM-based method used to approximate floodplain inundation by comparing terrain elevation to a specified water surface elevation (WSE). In this framework, the landscape is treated as if water fills the topography like a rising bathtub: all cells in the DEM with elevations lower than the assumed WSE are classified as inundated. To improve realism, the analysis typically restricts inundation to areas hydraulically connected to the river channel, which prevents isolated depressions from being incorrectly labeled as flooded. Because it relies only on elevation and an assumed stage, the bathtub method is computationally efficient and well suited for scenario screening, visualization, and communication of potential flood extents. However, it does not simulate hydraulic processes such as flow momentum, roughness, backwater effects, or levee overtopping, and therefore represents a static approximation rather than a physically dynamic flood model.

```{r}
# inputs
#dem_raw <- read_rds(here::here("data-raw", "DEM", "dem.rds"))  
# dem  <- read_rds(here::here("data-raw", "DEM", "dem_rast.rds"))  
dem_wrap <- terra::unwrap(read_rds(here::here("data-raw", "DEM", "dem_wrap.rds")))

river <- rivermile::sprague_river     
river_v <- vect(st_transform(river, crs(dem_wrap)))

# buffer river slightly so floodplain area included
river_buf <- buffer(river_v, 1000)  # meters

# crop DEM to buffer
dem_crop <- crop(dem_wrap, river_buf)
dem_crop <- mask(dem_crop, river_buf)

# now rasterize
river_r <- rasterize(river_v, dem_crop, field = 1)

# 2) choose a water surface elevation (example: 2.0 feet(?) above a reference river elevation)
# If you have gage WSE, use that directly (absolute elevation).
river_elev <- global(mask(dem_crop, river_r), "mean", na.rm = TRUE)[1,1]
wse <- river_elev + 2.0

# 3) initial inundation ("bathtub")
wet0 <- dem_crop <= wse 

# 4) keep only wet cells connected to the river
#    (seed = river cells; then expand through adjacent wet cells)
seed <- !is.na(river_r)
wet_connected <- terra::patches(wet0, directions=8)  # label connected wet patches
# find which patch IDs touch the river
touch_ids <- unique(values(wet_connected)[values(wet_connected) > 0 & values(seed)])
wet <- wet_connected %in% touch_ids

# 5) distance-from-channel across inundated area
dist_from_river <- distance(seed)  # planar distance in CRS units
dist_inundated <- mask(dist_from_river, wet)

# summary: max travel distance (in map units)
# global(dist_inundated, "max", na.rm = TRUE)

depth <- wse - dem_crop
depth <- mask(depth, wet, maskvalues = 0)   # keep inundated cells only
depth <- clamp(depth, lower = 0, values = TRUE)  # set negative depths to 0
depth[depth == 0] <- NA                      # hide zeros for cleaner map

depth_ll <- project(depth, "EPSG:4326", method = "bilinear")

vals <- values(depth_ll, mat = FALSE)
vals <- vals[is.finite(vals)]

```

### Sprague River Restoration Dataset

**data source**: https://data.riverscapes.net/rv/53f2c2a3-3324-4b34-94cc-2a3af2262c8f?do=1&lo=0&vm=VB_Access&bl=SAT&vp=-121.212807%7E42.421806%7E10.657

```{r}
off_channel_rearing <- st_read("../../klamathHabitatData/data-raw/PBR_Planning_Model_-_Bespoke-Sprague_River_PBR_Planning_Model/05_Restoration/02_Restoration_Space/Off_Channel_Rearing_Habitat.shp")

wetland_floodplain <- st_read("../../klamathHabitatData/data-raw/PBR_Planning_Model_-_Bespoke-Sprague_River_PBR_Planning_Model/05_Restoration/02_Restoration_Space/wetland_floodplain_corridor.shp")

forage_area <- st_read("../../klamathHabitatData/data-raw/PBR_Planning_Model_-_Bespoke-Sprague_River_PBR_Planning_Model/05_Restoration/02_Restoration_Space/Forage_Area.shp")


```

```{r}


pal <- colorNumeric(
  palette = "viridis",
  domain  = vals,
  na.color = "transparent"
)

wet_poly <- as.polygons(!is.na(depth_ll), dissolve = TRUE) |> st_as_sf()
bb <- st_bbox(wet_poly)

leaflet() |>
  addProviderTiles(providers$CartoDB.Positron, group = "Positron") |>
  addProviderTiles(providers$Esri.WorldImagery, group = "Imagery") |>
  addRasterImage(depth_ll, colors = pal, opacity = 0.65, group = "Depth") |>
  addPolygons(data = wetland_floodplain |> st_transform(4326),
              fillColor = "orange", fillOpacity = 0.5,
              color = "darkred", weight = 1, group = "Sprague RR Wetland") |> 
  addPolygons(data = off_channel_rearing |> st_transform(4326),
              fillColor = "brown", fillOpacity = 0.5,
              color = "darkgrey", weight = 1, group = "Sprague RR Off-Channel Rearing") |> 
  addPolygons(
    data = klamath_crops |> st_transform(4326),
    fillColor = "green", fillOpacity = 0.4,
    color = "darkgreen", weight = 1,
    group = "Crops"
  ) |>
  addPolylines(
    data = rivermile::all_klamath_rivers_line |> st_transform(4326),
    color = "blue", weight = 1.5, opacity = 0.5,
    label = ~river,
    labelOptions = labelOptions(
      noHide = TRUE, direction = "auto", textsize = "11px",
      style = list("color" = "navy", "font-weight" = "bold")
    ),
    group = "Rivers"
  ) |>
  addLegend(pal = pal, values = as.numeric(vals), title = "Water depth (meters)", opacity = 1) |>
  addLayersControl(
    baseGroups = c("Positron", "Imagery"),
    overlayGroups = c("Sprague RR Wetland", "Sprague RR Off-Channel Rearing", "Depth", "Crops", "Rivers"),
    options = layersControlOptions(collapsed = FALSE)
  ) |> 
  fitBounds(
    lng1 = as.numeric(bb["xmin"]),
    lat1 = as.numeric(bb["ymin"]),
    lng2 = as.numeric(bb["xmax"]),
    lat2 = as.numeric(bb["ymax"])
  )
```

### Compare methodolgoies 

The Sprague River resotoraton dataset includes wetland floodplain and I want to compare the total acreage of their analysis to the "Bath Tub" method. 

```{r}
sprague_acres <- wetland_floodplain |> 
  mutate(acres = area_sqm / 4046.86) |> # to acres 
  pull(acres)

```

```{r}
wet_cells <- global(!is.na(depth_ll), "sum", na.rm = TRUE)[1,1]
cell_area <- cellSize(depth_ll, unit = "m")   # SAFE for lat/long rasters
total_area_m2 <- global(cell_area * (!is.na(depth_ll)), "sum", na.rm = TRUE)[1,1]
bathtub_acres <- total_area_m2 / 4046.8564224
```

The Sprague Restoration analysis reported: `r sprague_acres` acres. 
The Bath Tub DEM approach reported: `r bathtub_acres` acres. 


```{r}
knitr::knit_exit()
```

## Wetlands 

Pre-process
```{r}
# choose a local projected CRS for distance/area (example: UTM)
target_crs <- "EPSG:26910"

#dem_p <- project(dem, target_crs)
#saveRDS(dem_p, here::here("data-raw", "DEM", "dem_p.rds"))
dem_p <- readRDS(here::here("data-raw", "DEM", "dem_p.rds"))
river_p <- st_transform(river, target_crs) |> vect()

# Limit to a corridor / floodplain to keep rasters small
river_buf <- buffer(river_p, 2000)          # tweak to your corridor width
dem_c <- crop(dem_p, river_buf) |> mask(river_buf)

# Optionally aggregate to 2–5 m for speed
# dem_c <- aggregate(dem_c, fact = 2)         # if original is 1 m -> 2 m


```

build inundation mask for activated space 

```{r eval=FALSE, include=FALSE}
make_inundation <- function(dem_r, river_geom, wse) {
  # river raster seed on DEM grid
  river_r <- rasterize(river_geom, dem_r, field = 1, background = NA)

  wet0 <- dem_r <= wse
  seed <- !is.na(river_r)

  # connected-component labeling of wet cells
  patch <- patches(wet0, directions = 8)

  # keep only patch IDs that touch the river seed
  ids <- unique(values(patch)[values(seed)])
  wet <- patch %in% ids

  list(wet = wet, river_r = river_r)
}

activated_acres <- function(wet_r, dem_r) {
  # wet_r is logical/0-1 on same grid as dem_r
  cell_area_m2 <- prod(res(dem_r))
  wet_cells <- global(wet_r, "sum", na.rm = TRUE)[1, 1]
  (wet_cells * cell_area_m2) / 4046.8564224
}


```

compute depth + activated area for a given WSE (scenario)

```{r eval=FALSE, include=FALSE}
one <- make_inundation(dem_c, river_p, wse = 1262.4)   # example WSE

wet   <- one$wet
depth <- (1262.4 - dem_c) |> mask(wet, maskvalues = 0) |> clamp(lower = 0, values = TRUE)
depth[depth == 0] <- NA

# activated space (area)
cell_area_m2 <- prod(res(dem_c))
activated_area_m2 <- global(wet, "sum", na.rm = TRUE)[1,1] * cell_area_m2
activated_area_ac  <- activated_area_m2 / 4046.8564224
activated_area_ac


```

Hyperperiod 

```{r}
# dummy wse_df
wse_df <- tibble(
  date = seq.Date(as.Date("2020-01-01"), as.Date("2020-12-31"), by = "day"),
  wse  = runif(length(date), min = 800, max = 1500)
)
```

```{r eval=FALSE, include=FALSE}
# wse_df: columns date, wse
# Keep the window you care about (example: Jan–May)
wse_jan_may <- wse_df |>
  mutate(date = as.Date(date)) |>
  filter(month(date) %in% 1:5) |>
  arrange(date)

wet_list <- lapply(wse_jan_may$wse, \(w) make_inundation(dem_c, river_p, w)$wet)
wet_stack <- rast(wet_list)
names(wet_stack) <- as.character(wse_jan_may$date)

# days inundated in the period
days_wet <- app(wet_stack, fun = sum, na.rm = TRUE)

# fraction of timesteps wet (0–1)
frac_wet <- days_wet / nlyr(wet_stack)

# activated space in the period: union across timesteps
activated_union <- app(wet_stack, fun = max, na.rm = TRUE)


```



```{r}

make_inundation <- function(dem_r, river_geom, wse) {
  river_r <- rasterize(river_geom, dem_r, field = 1, background = NA)

  wet0  <- dem_r <= wse
  seed  <- !is.na(river_r)
  patch <- patches(wet0, directions = 8)

  ids <- unique(values(patch)[values(seed)])
  patch %in% ids
}

activated_acres <- function(wet_r, dem_r) {
  cell_area_m2 <- prod(res(dem_r))
  wet_cells <- global(wet_r, "sum", na.rm = TRUE)[1, 1]
  (wet_cells * cell_area_m2) / 4046.8564224
}

activated_acres_by_month <- wse_df |>
  mutate(
    date  = as.Date(date),
    month = month(date)
  ) |>
  filter(!is.na(wse)) |>
  group_by(month, year = year(date)) |>
  rowwise() |>
  mutate(
    activated_acres = activated_acres(
      make_inundation(dem_c, river_p, wse),
      dem_c
    )
  ) |>
  ungroup()

activated_acres_by_month

```


