---
title: "Crop Report Exploration"
author: "Maddee Rubenson (FlowWest)"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 9, fig.height = 8)
library(tidyverse)
library(scales)

```

## Overview 

Crop Reports were provided from 1999 to 2019. Each report is a yearly requirement in the water contracts, all crop acreages are provided annually by irrigation districts. Reports include Oregon and California 

This document provides an overview of the contents of the reports for each year. 

### Crop Acreage and Value Stats 

#### Crop Statistics

**Water Rentals**: Contracts for the use of rental water are for one year's use; therefore, the area receiving this type of service

```{r include=FALSE}
crop_stats <- readxl::read_excel(here::here("data-raw", "crop_reports.xlsx"), sheet = "crop_statistics")
```

```{r echo=FALSE}
crop_stats |>
  filter(Category %in% c(
    "WATER RENTAL", 
    "IDLED LANDS", 
    "ROW CROP (Potatoes & Onions)", 
    "HAY", 
    "GRAIN", 
    "PASTURE"
  )) |>
  ggplot(aes(x = factor(Report), y = Value)) +
  geom_col(fill = "steelblue") +
  geom_text(
    aes(label = comma(Value)),
    vjust = -0.25,
    size = 3
  ) +
  facet_wrap(~ Category, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(
    x = "Year",
    y = "Acres",
    title = "Crop Statistics"
  ) +
  theme_minimal() 
```

#### Water Statistics (Acre-Feet of Water Delivered)

```{r include=FALSE}
water_stats <- readxl::read_excel(here::here("data-raw", "crop_reports.xlsx"), sheet = "water_statistics")
```

```{r echo=FALSE}
water_stats |>
  ggplot(aes(x = factor(Report), y = Value, fill = Category)) +
  geom_col(position = position_dodge()) +
  geom_text(
    aes(label = comma(Value)),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  xlab("Year") +
  ylab("AF–Water Delivered") +
  ggtitle("Water Statistics (AF Delivered)") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )

```

### Acreage Summary

#### Lands In and Out of Irrigation

```{r include=FALSE}
acreage_summary <- readxl::read_excel(here::here("data-raw", "crop_reports.xlsx"), sheet = "acreage_summary")
```

```{r echo=FALSE}
# CALIFORNIA
acreage_summary |>
  filter(!grepl("TOTAL", Description),
         Description !=  "ACRES IRRIGATED (LINES 11 + 12)", 
         Description != 'ACRES IRRIGATED (LINES 11 +12)')  |>
  filter(State == "CA") |> 
  filter(!is.na(Total)) |> 
  mutate(in_irrigation = ifelse(in_irrigation == "yes", "In Irrigation", "Not in Irrigation")) |> 
  ggplot(aes(x = factor(Report), y = Total, fill = Description)) +
  geom_col(position = position_dodge()) +
  geom_text(
    aes(label = comma(Total)),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  facet_wrap(~in_irrigation, scales = "free_y") + 
  xlab("Year") +
  ylab("Acres") +
  ggtitle("California: Acreage Summary for Lands in and out of Irrigation Rotation") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma) +
  guides(
    fill = guide_legend(
      ncol   = 2,    
      byrow  = TRUE  
    )
  ) +
  theme_minimal() +
  theme(
    legend.position   = "top",
    legend.title      = element_blank(),
    legend.text       = element_text(size = 8),             
    legend.key.width  = unit(1.2, "cm"),                    
    strip.text        = element_text(size = 10),           
    axis.text.x       = element_text(angle = 45, hjust = 1) 
  )

# OREGON
acreage_summary |>
  filter(!grepl("TOTAL", Description),
         Description !=  "ACRES IRRIGATED (LINES 11 + 12)", 
         Description != 'ACRES IRRIGATED (LINES 11 +12)')  |>
  filter(State == "OR") |> 
  filter(!is.na(Total)) |> 
  mutate(in_irrigation = ifelse(in_irrigation == "yes", "In Irrigation", "Not in Irrigation")) |> 
  ggplot(aes(x = factor(Report), y = Total, fill = Description)) +
  geom_col(position = position_dodge()) +
  geom_text(
    aes(label = comma(Total)),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  facet_wrap(~in_irrigation, scales = "free_y") + 
  xlab("Year") +
  ylab("Acres") +
  ggtitle("Oregon: Acreage Summary for Lands in and out of Irrigation Rotation") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma) +
  guides(
    fill = guide_legend(
      ncol   = 2,    
      byrow  = TRUE  
    )
  ) +
  theme_minimal() +
  theme(
    legend.position   = "top",
    legend.title      = element_blank(),
    legend.text       = element_text(size = 8),             
    legend.key.width  = unit(1.2, "cm"),                    
    strip.text        = element_text(size = 10),           
    axis.text.x       = element_text(angle = 45, hjust = 1) 
  )

```

#### Number of Farms

```{r include=FALSE}
no_farms <- readxl::read_excel(here::here("data-raw", "crop_reports.xlsx"), sheet = "number_of_farms_and_pop") |> 
  janitor::clean_names()
```

```{r echo=FALSE}
no_farms |> 
  filter(grepl('FARMS', number_of_farms_and_population)) |> 
  mutate(
    number_of_farms_and_population = str_trim(str_remove(number_of_farms_and_population, "^(41|42)\\s*"))
  ) |> 
  ggplot(aes(x = factor(report), y = no_of_farms, fill = number_of_farms_and_population)) +
  geom_col(position = position_dodge()) +
  facet_wrap(~state) + 
  geom_text(
    aes(label = comma(no_of_farms)),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  xlab("Year") +
  ylab("Number of Farms in Oregon and California (est)") +
  ggtitle("Estimated Number of Farms") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )
```

### Crop Production

```{r include=FALSE}
crops <- readxl::read_excel(here::here("data-raw", "crop_reports.xlsx"), sheet = "crop_production") |> 
  janitor::clean_names()
```

#### Total Yield

Where *total yield = acres x per acre yield*

Note that crop categories don't necessarily have consistent units.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# CALIFORNIA
crops |> 
  group_by(crop_category, report, state) |> 
  summarise(total_yield = sum(total_yield, na.rm = T),
            total_value = sum(total_value, na.rm = T)) |> 
  filter(state == "CA") |> 
  ggplot(aes(x = factor(report), y = total_yield, fill = crop_category)) +
  geom_col(position = position_dodge()) +
  geom_text(
    aes(label = comma(total_yield)),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  xlab("Year") +
  ylab("Total Yield") +
  ggtitle("California: Total Yield by Crop Type") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )

# OREGON
crops |> 
  group_by(crop_category, report, state) |> 
  summarise(total_yield = sum(total_yield, na.rm = T),
            total_value = sum(total_value, na.rm = T)) |> 
  filter(state == "OR") |> 
  ggplot(aes(x = factor(report), y = total_yield, fill = crop_category)) +
  geom_col(position = position_dodge()) +
  geom_text(
    aes(label = comma(total_yield)),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  xlab("Year") +
  ylab("Total Yield") +
  ggtitle("Oregon: Total Yield by Crop Type") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )
```

#### Total Value of Crops

Where *total value = acres x per acre crop value*

```{r echo=FALSE, message=FALSE, warning=FALSE}
# CALIFORNIA
crops |> 
  group_by(crop_category, report, state) |> 
  summarise(total_yield = sum(total_yield, na.rm = T),
            total_value = sum(total_value, na.rm = T)) |> 
  filter(state == "CA") |> 
  ggplot(aes(x = factor(report), y = total_value, fill = crop_category)) +
  geom_col(position = position_dodge()) +
  geom_text(
    aes(label = comma(total_value)),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  xlab("Year") +
  ylab("Total Value") +
  ggtitle("California: Total Value by Crop Type") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )

# OREGON
crops |> 
  group_by(crop_category, report, state) |> 
  summarise(total_yield = sum(total_yield, na.rm = T),
            total_value = sum(total_value, na.rm = T)) |> 
  filter(state == "OR") |> 
  ggplot(aes(x = factor(report), y = total_value, fill = crop_category)) +
  geom_col(position = position_dodge()) +
  geom_text(
    aes(label = comma(total_value)),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  xlab("Year") +
  ylab("Total Value") +
  ggtitle("Oregon: Total Value by Crop Type") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )
```

### Crop Stats

https://www.usbr.gov/projects/pdf.php?id=7

focus on getting the water information
