---
title: "Untitled"
author: "Maddee Rubenson (FlowWest)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(sf)
library(tidyverse)
library(leaflet)
library(terra)

```

```{r include=FALSE}
klamath_basin <- st_read('klamath_basin_outline/R8_FAC_Klamath_Basin_WFL1.shp')
ggplot() + 
  geom_sf(data = klamath_basin)
```

```{r include=FALSE}
or_field <- st_layers(here::here("data-raw", "cropET", "Appendix_6a_Field_and_HUC_Consumptive_Use_Geodatabase", "or_field_geopackage.gpkg"))
layers <- unique(or_field$name)

crop <- st_read(
  here::here("data-raw", "cropET", "Appendix_6a_Field_and_HUC_Consumptive_Use_Geodatabase", "or_field_geopackage.gpkg"),
  layer = "CROP"
) |> 
  st_transform(st_crs(klamath_basin))

klamath_crops <- sf::st_intersection(crop, klamath_basin)

klamath_crops |> glimpse()

klamath_crops |> 
  ggplot() + 
  geom_sf()
```

```{r include=FALSE}
unique(klamath_crops$ITYPE)

unique(klamath_crops$CROP_2022)

```

## Levees

**Data source: ** https://resilience.climate.gov/maps/87acff1ba86c40098b59472292de3d11/about

```{r}

url <- "https://services2.arcgis.com/FiaPA4ga0iQKduv3/ArcGIS/rest/services/NLD2_PUBLIC_v1/FeatureServer/0/query?where=1%3D1&outFields=*&f=geojson"

nld <- st_read(url)

nld_klamath <- nld |> 
  st_transform(st_crs(klamath_basin)) |> 
  sf::st_intersection(klamath_basin) |> 
  janitor::clean_names()

ggplot() + 
  geom_sf(data = klamath_basin) + 
  geom_sf(data = nld_klamath) 
```
```{r}
leaflet() |>
  addProviderTiles(providers$CartoDB.Positron, group = "Light") |>
  addProviderTiles(providers$Esri.WorldImagery, group = "Imagery") |>
  addPolygons(
    data = klamath_crops |> st_transform(4326),
    fillColor = "green",
    fillOpacity = 0.4,
    color = "darkgreen",
    weight = 1,
    group = "Crops"
  ) |>
  addPolylines(
    data = rivermile::all_klamath_rivers_line,
    color = "blue",
    weight = 1.5,
    opacity = 0.5,  
    label = ~river, 
    labelOptions = labelOptions(
      noHide = TRUE,         
      direction = "auto",
      textsize = "11px",
      style = list(
        "color" = "navy",
        "font-weight" = "bold"
      )
    ),
    group = "Rivers"
  ) |>
  addLayersControl(
    baseGroups = c("Light", "Imagery"),
    overlayGroups = c("Crops", "Rivers"),
    options = layersControlOptions(collapsed = FALSE)
  )
  

```

## DEM

Download the DEM using the `elevatr` package

DEM is in meters.

We have access to a high resolution DEM and bathymetric dataset if we want to continue this route. The following is exploratory. 

```{r eval=FALSE, include=FALSE}
library(elevatr)
library(terra)

aoi <- klamath_basin

dem <- get_elev_raster(
  locations = aoi,
  z = 12,          # higher = finer (~10–30 m typical)
  clip = "locations"
)

dem_rast <- rast(dem)
plot(dem_rast)

saveRDS(dem, here::here("data-raw", "DEM", "dem.rds"))
saveRDS(dem_rast, here::here("data-raw", "DEM", "dem_rast.rds"))

```
## Map Floodplain Inundation 

The bathtub approach is a simple DEM-based method used to approximate floodplain inundation by comparing terrain elevation to a specified water surface elevation (WSE). In this framework, the landscape is treated as if water fills the topography like a rising bathtub: all cells in the DEM with elevations lower than the assumed WSE are classified as inundated. To improve realism, the analysis typically restricts inundation to areas hydraulically connected to the river channel, which prevents isolated depressions from being incorrectly labeled as flooded. Because it relies only on elevation and an assumed stage, the bathtub method is computationally efficient and well suited for scenario screening, visualization, and communication of potential flood extents. However, it does not simulate hydraulic processes such as flow momentum, roughness, backwater effects, or levee overtopping, and therefore represents a static approximation rather than a physically dynamic flood model.

```{r}
# inputs
#dem_raw <- readRDS(here::here("data-raw", "DEM", "dem.rds"))  
dem  <- readRDS(here::here("data-raw", "DEM", "dem_rast.rds"))            
river <- rivermile::sprague_river     
river_v <- vect(st_transform(river, crs(dem)))

# buffer river slightly so floodplain area included
river_buf <- buffer(river_v, 1000)  # meters? 

# crop DEM to buffer
dem_crop <- crop(dem, river_buf)
dem_crop <- mask(dem_crop, river_buf)

# now rasterize
river_r <- rasterize(river_v, dem_crop, field = 1)

# 2) choose a water surface elevation (example: 2.0 feet(?) above a reference river elevation)
# If you have gage WSE, use that directly (absolute elevation).
river_elev <- global(mask(dem_crop, river_r), "mean", na.rm = TRUE)[1,1]
wse <- river_elev + 2.0

# 3) initial inundation ("bathtub")
wet0 <- dem_crop <= wse 

# 4) keep only wet cells connected to the river
#    (seed = river cells; then expand through adjacent wet cells)
seed <- !is.na(river_r)
wet_connected <- terra::patches(wet0, directions=8)  # label connected wet patches
# find which patch IDs touch the river
touch_ids <- unique(values(wet_connected)[values(wet_connected) > 0 & values(seed)])
wet <- wet_connected %in% touch_ids

# 5) distance-from-channel across inundated area
dist_from_river <- distance(seed)  # planar distance in CRS units
dist_inundated <- mask(dist_from_river, wet)

# summary: max travel distance (in map units)
global(dist_inundated, "max", na.rm = TRUE)

```
```{r}
# --- Multidirectional hillshade from dem_crop ---
slope  <- terrain(dem_crop, v = "slope",  unit = "radians")
aspect <- terrain(dem_crop, v = "aspect", unit = "radians")

# pick a few azimuths (degrees); altitude ~ 45° is a common default
az <- c(315, 45, 90, 135, 180, 225, 270)

hs_stack <- rast(lapply(az, \(a) shade(slope, aspect, angle = 45, direction = a)))
hs_multi <- app(hs_stack, fun = mean, na.rm = TRUE)

# (optional) stretch contrast a bit (simple linear stretch)
hs_multi <- clamp(hs_multi, lower = quantile(values(hs_multi), 0.02, na.rm = TRUE),
                            upper = quantile(values(hs_multi), 0.98, na.rm = TRUE),
                            values = TRUE)

# project to lon/lat for leaflet
hs_ll <- project(hs_multi, "EPSG:4326", method = "bilinear")

# grayscale palette
hs_vals <- values(hs_ll, mat = FALSE)
hs_vals <- hs_vals[is.finite(hs_vals)]
hs_pal <- colorNumeric(gray.colors(256), domain = hs_vals, na.color = "transparent")

```


```{r}
depth <- wse - dem_crop
depth <- mask(depth, wet, maskvalues = 0)   # keep inundated cells only
depth <- clamp(depth, lower = 0, values = TRUE)  # set negative depths to 0
depth[depth == 0] <- NA                      # hide zeros for cleaner map

depth_ll <- project(depth, "EPSG:4326", method = "bilinear")

vals <- values(depth_ll, mat = FALSE)
vals <- vals[is.finite(vals)]

pal <- colorNumeric(
  palette = "viridis",
  domain  = vals,
  na.color = "transparent"
)

wet_poly <- as.polygons(!is.na(depth_ll), dissolve = TRUE) |> st_as_sf()
bb <- st_bbox(wet_poly)

leaflet() |>
  addProviderTiles(providers$CartoDB.Positron, group = "Positron") |>
  addProviderTiles(providers$Esri.WorldImagery, group = "Imagery") |>
    addRasterImage(hs_ll, colors = hs_pal, opacity = 0.8, group = "Hillshade") |>
  addRasterImage(depth_ll, colors = pal, opacity = 0.65, group = "Depth") |>
  addPolygons(
    data = klamath_crops |> st_transform(4326),
    fillColor = "green", fillOpacity = 0.4,
    color = "darkgreen", weight = 1,
    group = "Crops"
  ) |>
  addPolylines(
    data = rivermile::all_klamath_rivers_line |> st_transform(4326),
    color = "blue", weight = 1.5, opacity = 0.5,
    label = ~river,
    labelOptions = labelOptions(
      noHide = TRUE, direction = "auto", textsize = "11px",
      style = list("color" = "navy", "font-weight" = "bold")
    ),
    group = "Rivers"
  ) |>
  addLegend(pal = pal, values = as.numeric(vals), title = "Water depth (meters)", opacity = 1) |>
  addLayersControl(
    baseGroups = c("Positron", "Imagery"),
    overlayGroups = c("Hillshade", "Depth", "Crops", "Rivers"),
    options = layersControlOptions(collapsed = FALSE)
  ) |> 
  fitBounds(
    lng1 = as.numeric(bb["xmin"]),
    lat1 = as.numeric(bb["ymin"]),
    lng2 = as.numeric(bb["xmax"]),
    lat2 = as.numeric(bb["ymax"])
  )
```
